---
title: "30-day mortality"
author: "University of Waterloo Health Data Science Lab"
date: "March 2018"
output: pdf_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/path to project/")
```

```{r libraries, include=FALSE}
library(data.table)
library(rms)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(xtable)
library(caret)
library(MLmetrics)
library(PRROC)
library(pROC)
library(questionr)
```

# Data preparation

```{r main}
set.seed(4375) # "HDSL"

admissions_df <- read.csv("Data/patients_df.csv")
sapsii_df <- read.csv("Data/sapsii_df.csv")
notes_df <- fread("Data/notes_df_sntmnt.csv")
mimicii_patients_df <- read.csv("Data/mimic2_patients_df.csv")
mimicii_admissions_df <- read.csv("Data/mimic2_admissions_df.csv")

notes_df$text <- NULL

notes_df <- notes_df %>% left_join(admissions_df, by="hadm_id")

# Get the difference between the time of death and the time that the note was taken.
# We don't want to look at notes close to death (~12hrs)

deathtime.POSIX <- as.POSIXlt(strptime(notes_df$dod, "%Y-%m-%d %H:%M:%S"))
charttime.POSIX <- as.POSIXlt(strptime(notes_df$charttime, "%Y-%m-%d %H:%M:%S"))
notes_df$death_chart_diffhrs <- difftime(deathtime.POSIX, 
                                         charttime.POSIX, 
                                         units="hours")

setDT(notes_df)

notes_df <- notes_df[is.na(death_chart_diffhrs) | 
                       death_chart_diffhrs > 12]

notes_df <- notes_df[is.na(iserror)]

analysis_df <- admissions_df

setDT(analysis_df) 

analysis_df[dod != "", 
            mortality_days := difftime(as.Date(dod), 
                                       as.Date(admittime), units="days")]

# Remove patients with data entry error: death time recorded before admission.
analysis_df <- analysis_df[is.na(mortality_days) | mortality_days >= 0]

analysis_df[!is.na(mortality_days) & mortality_days < 30, mortality_30d := 1]
analysis_df[is.na(mortality_30d), mortality_30d := 0]
analysis_df$mortality_30d <- factor(analysis_df$mortality_30d)

```

```{r main2}

setDT(analysis_df) 
analysis_df$first_careunit <- factor(analysis_df$first_careunit)

### For counting number of notes per patient

mytbl <- notes_df
nrow(mytbl)
length(unique(mytbl$hadm_id))

count.tbl <- count(mytbl, hadm_id)
count.tbl

analysis_df_counted <- left_join(analysis_df, count.tbl, by="hadm_id")
setDT(analysis_df_counted)
analysis_df_counted[is.na(n), n := 0]


mean(analysis_df_counted$n)
sd(analysis_df_counted$n)
median(analysis_df_counted$n)
summary(analysis_df_counted$n)


a_df_counted_nonzeronotes <- analysis_df_counted[n > 0]
nrow(a_df_counted_nonzeronotes)
mean(a_df_counted_nonzeronotes$n)
sd(a_df_counted_nonzeronotes$n)
summary(a_df_counted_nonzeronotes$n)

hist(a_df_counted_nonzeronotes$n)

# Break down by 30-day mortality group
# Survived
a_df_counted_nonzeronotes <- analysis_df_counted[n > 0 & mortality_30d == 0]
nrow(a_df_counted_nonzeronotes)
mean(a_df_counted_nonzeronotes$n)
sd(a_df_counted_nonzeronotes$n)
summary(a_df_counted_nonzeronotes$n)

# Expired
a_df_counted_nonzeronotes <- analysis_df_counted[n > 0 & mortality_30d == 1]
nrow(a_df_counted_nonzeronotes)
mean(a_df_counted_nonzeronotes$n)
sd(a_df_counted_nonzeronotes$n)
summary(a_df_counted_nonzeronotes$n)


### End block

notes_df <- notes_df %>% 
  group_by(hadm_id) %>% 
  summarize(mean_polarity = mean(polarity),
            mean_subjectivity = mean(subjectivity))

# joining notes by admission id.
analysis_df <- analysis_df %>% left_join(notes_df, by="hadm_id", copy=TRUE)

# joining SAPS II and SOFA scores by ICU stay ID.
analysis_df <- analysis_df %>% left_join(sapsii_df, by="icustay_id", copy=TRUE)
```

Notice there are some data where patients are recorded to be very old (~300 yrs old). This was done by MIMIC to anonymize patients over the age of 89.

However, in MIMIC-II, the ages of patients were recorded after death. Therefore, we will randomly sample ages from the >89 group from MIMIC-II and assign their ages to the >89 group from MIMIC-III. We will do so by gender/sex since as we will see, their age distributions are slightly different.

```{r}
mimicii_age_dataset <- mimicii_admissions_df %>%
  group_by(subject_id) %>%
  arrange(admit_dt) %>%
  filter(row_number()==1) 

mimicii_age_dataset <- mimicii_age_dataset %>% 
  left_join(mimicii_patients_df, by="subject_id")

mimicii_age_dataset$first_admit_age <- as.numeric(as.Date(mimicii_age_dataset$admit_dt) - 
                                        as.Date(mimicii_age_dataset$dob)) / 365.242

setDT(mimicii_age_dataset)
setDT(analysis_df)

ggplot(mimicii_age_dataset[first_admit_age > 89 & first_admit_age < 120]) + 
  geom_histogram(aes(first_admit_age, fill=sex), bins=20, alpha=0.6) + 
  ggtitle("Histogram of age of MIMICII patients by sex")

analysis_df[gender == "M" & first_admit_age > 89]$first_admit_age <-
  sample(mimicii_age_dataset[sex == "M" & first_admit_age > 89 & first_admit_age < 120]$first_admit_age,
         size = nrow(analysis_df[gender == "M" & first_admit_age > 89]),
         replace = TRUE)

analysis_df[gender == "F" & first_admit_age > 89]$first_admit_age <-
  sample(mimicii_age_dataset[sex == "F" & first_admit_age > 89 & first_admit_age < 120]$first_admit_age,
         size = nrow(analysis_df[gender == "F" & first_admit_age > 89]),
         replace = TRUE)

analysis_df_w_pol <- analysis_df %>%
  filter(!is.na(mean_polarity)) 
```

Code for producting summary statistics in table 2

```{r}

adf.survived <- analysis_df_w_pol[analysis_df_w_pol$mortality_30d == 0,]
adf.expired <- analysis_df_w_pol[analysis_df_w_pol$mortality_30d == 1,]

sum(adf.survived$first_careunit == "CCU") / nrow(adf.survived)
sum(adf.expired$first_careunit == "CCU") / nrow(adf.expired)

sum(adf.survived$first_careunit == "CSRU") / nrow(adf.survived)
sum(adf.expired$first_careunit == "CSRU") / nrow(adf.expired)

sum(adf.survived$first_careunit == "MICU") / nrow(adf.survived)
sum(adf.expired$first_careunit == "MICU") / nrow(adf.expired)

sum(adf.survived$first_careunit == "SICU") / nrow(adf.survived)
sum(adf.expired$first_careunit == "SICU") / nrow(adf.expired)

sum(adf.survived$first_careunit == "TSICU") / nrow(adf.survived)
sum(adf.expired$first_careunit == "TSICU") / nrow(adf.expired)

mean(adf.survived$sapsii)
sd(adf.survived$sapsii)
mean(adf.expired$sapsii)
sd(adf.expired$sapsii)

mean(adf.survived$first_admit_age)
sd(adf.survived$first_admit_age)
mean(adf.expired$first_admit_age)
sd(adf.expired$first_admit_age)

sum(adf.survived$gender == "F") / nrow(adf.survived)
sum(adf.expired$gender == "F") / nrow(adf.expired)

mean(adf.survived$mean_polarity)
mean(adf.expired$mean_polarity)

mean(adf.survived$mean_subjectivity)
mean(adf.expired$mean_subjectivity)

t.test(adf.expired$mean_polarity, adf.survived$mean_polarity)
t.test(adf.expired$mean_subjectivity, adf.survived$mean_subjectivity)

spearman.test(analysis_df_w_pol$mean_polarity, analysis_df_w_pol$sapsii)
spearman.test(analysis_df_w_pol$mean_subjectivity, analysis_df_w_pol$sapsii)
```

# Fit logistic regression model and plot results

```{r}

analysis_df_w_pol$mean_polarity <- 10 * analysis_df_w_pol$mean_polarity
analysis_df_w_pol$mean_subjectivity <- 10 * analysis_df_w_pol$mean_subjectivity

gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}

model_mortality_30d = glm(mortality_30d ~ mean_polarity + 
                            + mean_subjectivity + sapsii + 
                            first_careunit + gender, 
                          family=binomial(link='logit'), 
                          data=analysis_df_w_pol)

summary(model_mortality_30d)

model.odds.ratio <- odds.ratio(model_mortality_30d)

model.odds.ratio

lrm_30d <- lrm(model_mortality_30d$formula, data = analysis_df_w_pol)

# Calculate Nagelkerke pseudo-R^2
lrm_30d$stats["R2"]

output.folder <- "~/path to project/results"

analysis_df_w_pol$mean_polarity <- 0.1 * analysis_df_w_pol$mean_polarity
analysis_df_w_pol$mean_subjectivity <- 0.1 * analysis_df_w_pol$mean_subjectivity

pol.vs.mort.plot <- ggplot(analysis_df_w_pol, aes(mean_polarity, mortality_30d))
setDF(analysis_df_w_pol)
mean_mean_polarity_df <- 
  data.frame(mortality_30d = c("expired", "survived"), 
             mean_mean_polarity = 
               c(mean(analysis_df_w_pol[analysis_df_w_pol$mortality_30d == 1, "mean_polarity"]),
                 mean(analysis_df_w_pol[analysis_df_w_pol$mortality_30d == 0, "mean_polarity"])))

jitter.plot.mp <- pol.vs.mort.plot + geom_jitter(size = 0.005, alpha=1) +
  xlab("Mean polarity") + 
  scale_y_discrete("30-day mortality", labels = c("0" = "survived", "1" = "expired")) +
  geom_vline(aes(xintercept = mean_mean_polarity, colour = mortality_30d), data=mean_mean_polarity_df) +
  scale_colour_discrete(name = "30-day mortality") + theme(legend.position = "none", text=element_text(size=8))

sub.vs.mort.plot <- ggplot(analysis_df_w_pol, aes(mean_subjectivity, mortality_30d))
mean_mean_subjectivity_df <-
  data.frame(mortality_30d = c("expired", "survived"),
             mean_mean_subjectivity = 
               c(mean(analysis_df_w_pol[analysis_df_w_pol$mortality_30d == 1, "mean_subjectivity"]),
                 mean(analysis_df_w_pol[analysis_df_w_pol$mortality_30d == 0, "mean_subjectivity"])))

jitter.plot.ms <- sub.vs.mort.plot + geom_jitter(size = 0.005, alpha = 1) + 
  xlab("Mean subjectivity") + 
  scale_y_discrete("30-day mortality", labels = c("0" = "survived", "1" = "expired")) +
  geom_vline(aes(xintercept = mean_mean_subjectivity, colour = mortality_30d), data=mean_mean_subjectivity_df) +
  scale_colour_discrete(name = "30-day mortality") + 
  theme(text=element_text(size=8), axis.title.y = element_blank(), 
        axis.text.y = element_blank(), axis.ticks.y = element_blank())

jitter.plot <- grid.arrange(jitter.plot.mp, jitter.plot.ms, ncol = 2)

jitter.plot

ggsave(file.path(output.folder, "jitter_plot.tiff"), 
       plot = jitter.plot, width=150, height=84, units="mm", dpi=300)

# Mixture histograms

setDT(analysis_df_w_pol)
levels(analysis_df_w_pol$mortality_30d) <- c("survived", "expired")
pol.mixture.histogram <- ggplot() +
  geom_histogram(aes(y = ..density.., x = mean_polarity, fill=mortality_30d), 
                 alpha = 0.3, bins = 60, data=analysis_df_w_pol[mortality_30d == "expired"]) +
  geom_histogram(aes(y = ..density.. , x = mean_polarity, fill=mortality_30d), 
                 alpha = 0.3, bins = 60, data=analysis_df_w_pol[mortality_30d == "survived"]) +
  ylab("Density") + xlab("Mean sentiment polarity") +
  geom_vline(aes(xintercept = mean_mean_polarity, colour = mortality_30d), data=mean_mean_polarity_df) + 
  scale_fill_manual(values = gg_color_hue(2),
                    name="30-day\nmortality") +
  scale_colour_manual(values = gg_color_hue(2), name="30-day\nmortality") + 
  theme(text = element_text(size = 8), legend.position = "none")

sub.mixture.histogram <- ggplot() +
  geom_histogram(aes(y = ..density.., x = mean_subjectivity, fill=mortality_30d), 
                 alpha = 0.3, bins = 60, data=analysis_df_w_pol[mortality_30d == "expired"]) +
  geom_histogram(aes(y = ..density.. , x = mean_subjectivity, fill=mortality_30d), 
                 alpha = 0.3, bins = 60, data=analysis_df_w_pol[mortality_30d == "survived"]) +
  ylab("Density") + xlab("Mean sentiment subjectivity") +
  geom_vline(aes(xintercept = mean_mean_subjectivity, colour = mortality_30d), data=mean_mean_subjectivity_df) + 
  scale_fill_manual(values = gg_color_hue(2),
                    name="30-day\nmortality") +
  scale_colour_manual(values = gg_color_hue(2), name="30-day\nmortality") + 
  theme(text=element_text(size=8), axis.title.y = element_blank(), 
        axis.text.y = element_blank(), axis.ticks.y = element_blank())

mixture.histogram <- grid.arrange(pol.mixture.histogram, sub.mixture.histogram, ncol = 2)

mixture.histogram

ggsave(file.path(output.folder,"mixture_histogram.tiff"), plot = mixture.histogram, 
       width=150, height=84, units="mm", dpi=300)

```

# Cross-validation and ROC

```{r}
# Reset data to where it was before modifying for plots
analysis_df_w_pol <- analysis_df %>%
  filter(!is.na(mean_polarity))

set.seed(4375)

# Change order of factors (required for using prSummary)
analysis_df_w_pol$mortality_30d <- factor(analysis_df_w_pol$mortality_30d, levels = c("1", "0"))
levels(analysis_df_w_pol$mortality_30d) <- c("expired", "survived")

train_control <- trainControl(method="repeatedcv", number=10, repeats=50,
                              summaryFunction = twoClassSummary, classProbs = TRUE, savePredictions = FALSE)

roc.model1 <- train(mortality_30d ~ 
                     mean_polarity + mean_subjectivity + sapsii + first_careunit + gender, 
                   data=analysis_df_w_pol, 
                   method="glm", family=binomial(), trControl=train_control)

roc.model2 <- train(mortality_30d ~ 
                     sapsii + first_careunit + gender, 
                   data=analysis_df_w_pol, 
                   method="glm", family=binomial(), trControl=train_control)

roc.model1$results
roc.model2$results


### Precision recall ###

train_control <- trainControl(method="repeatedcv", number=10, repeats=50,
                              summaryFunction = prSummary, classProbs = TRUE, savePredictions = FALSE)

roc.model1 <- train(mortality_30d ~ 
                     mean_polarity + mean_subjectivity + sapsii + first_careunit + gender, 
                   data=analysis_df_w_pol, 
                   method="glm", family=binomial(), trControl=train_control)

roc.model2 <- train(mortality_30d ~ 
                     sapsii + first_careunit + gender, 
                   data=analysis_df_w_pol, 
                   method="glm", family=binomial(), trControl=train_control)

roc.model1$results
roc.model2$results
```

### Bootstrap CI's
```{r}
### Paired method ###

# Reset data to where it was before modifying for plots. 
# This is technically reduntant if we run the AUC's first, but 
# doesn't hurt, and will allow users to jump over the previous AUC section.
analysis_df_w_pol <- analysis_df %>%
  filter(!is.na(mean_polarity))

analysis_df_w_pol$mortality_30d <- factor(analysis_df_w_pol$mortality_30d, levels = c("1", "0"))
levels(analysis_df_w_pol$mortality_30d) <- c("expired", "survived")

train_control <- trainControl(method="repeatedcv", number=10, repeats=1,
                              summaryFunction = twoClassSummary, classProbs = TRUE, savePredictions = TRUE)
seed <- 4375
set.seed(seed)
roc.model1 <- train(mortality_30d ~ 
                      mean_polarity + mean_subjectivity + sapsii + first_careunit + gender, 
                    data=analysis_df_w_pol, 
                    method="glm", family=binomial(), trControl=train_control)
set.seed(seed)
roc.model2 <- train(mortality_30d ~ 
                      sapsii + first_careunit + gender, 
                    data=analysis_df_w_pol, 
                    method="glm", family=binomial(), trControl=train_control)

library(precrec)
n.boot <- 2000 # Use 2000 for paper

reduced.data <- analysis_df_w_pol[c("mean_polarity", 
                                    "sapsii", "first_careunit", 
                                    "gender", "mortality_30d")]

auc.df <- roc.model1$pred
auc.df$expired1 <- roc.model1$pred$expired
auc.df$expired2 <- roc.model2$pred$expired

auc.df <- auc.df[c("obs", "expired1", "expired2")]

for (j in c(1,2))
{
  auc.list <- lapply(1:n.boot, function(i){
  
    data.sample <- sample_n(auc.df, nrow(auc.df), replace = TRUE)
    auc.1 <- evalmod(labels = (data.sample$obs == "expired"), scores = data.sample$expired1)
    auc.2 <- evalmod(labels = (data.sample$obs == "expired"), scores = data.sample$expired2)
    
    auc.1 <- attr(auc.1, "auc")[j,4] # use [1,4] for ROC, and [2,4] for PRC
    auc.2 <- attr(auc.2, "auc")[j,4]
    
    auc.diff <- auc.1 - auc.2
    
    return(auc.diff)
  }) 
  
  auc.results <- do.call(rbind, auc.list)

  if(j == 1)
  {
    print("AUROC difference bootstrap 95% CI:")
  }
  else 
  {
    print("AUPRC difference bootstrap 95% CI:")
  }
  
  print(quantile(auc.results, c(0.025, 0.975)))
}
```

